---
title: "Supplementary Material: Bayesian estimation in scale mixture of skew-normal linear mixed models using Hamiltonian Monte Carlo"
author: "Fernanda L Schumacher, Larissa A Matos, Pedro L Ramos, and   Francisco Louzada"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
    theme: united
---

```{r setup, include=F, echo=T, results='hide'}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
library(tictoc)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(loo)
library(rstan)
library(posterior)
library(bayesplot)
library(HDInterval)
library(skewlmm)
library(bayesplot)

# Stan optimization
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Introduction

This document provides the full reproducible workflow for the simulation and estimation of the AR(1) Skew-t Linear Mixed Model (AR1-ST-LMM) discussed in the main text, in addition for comparisons with a Gaussian model and a brief sensitivity analysis for the degrees of freedom parameter $\nu$. All materials needed to reproduce this code are available at \url{https://github.com/fernandalschumacher/HMC-SMSN-LMM}.

Stan is required for running this code. For information on installing Stan, please check
\url{https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started#installation-of-rstan}.

## Loading packages

```{r packages, eval=F}
library(tictoc)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(loo)
library(rstan)
library(posterior)
library(bayesplot)
library(HDInterval)
library(skewlmm)
library(bayesplot)
```


## Data Generation

We generate a longitudinal dataset with $m=200$ subjects and $n_j=5$ observations per subject from an AR(1)-ST-LMM using the skewlmm package.

```{r datgen}
set.seed(955)
nj1 <- 5 
m <- 200 
D1_true <- matrix(c(.5,.1,.1,1), ncol=2) 

gendatList <- map(rep(nj1, m), function(nj) {
  skewlmm::rsmsn.lmm(1:nj, cbind(1, 1:nj), cbind(1, 1:nj), 
                     sigma2 = .25, D1 = D1_true, beta = c(1, 2), 
                     lambda = c(1,3), depStruct = "ARp", 
                     phi = .5, distr = 'st', nu = 6)
})

gendat <- bind_rows(gendatList, .id = "ind") %>%
  mutate(indnum = as.numeric(ind)) %>%
  arrange(indnum)

# Visualization
ggplot(gendat, aes(x = x, y = y, group = ind)) + 
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = 1), geom = "line", fun = mean, col = "blue", size = 1.5) +
  theme_minimal() + 
  labs(title = "Simulated Individual Trajectories", x = "Time", y = "Response")
```


## AR(1)-ST-LMM Fitting using Stan

### Setup Data for Stan
```{r datinfo}
N <- nrow(gendat)
n <- n_distinct(gendat$ind)
x_mat <- model.matrix(y ~ x, data = gendat)
z_mat <- model.matrix(~ x, data = gendat)
l_cols <- ncol(x_mat)
q1_cols <- ncol(z_mat)
njvec <- as.numeric(table(gendat$indnum))

data_list_stan <- list(
  y = gendat$y, x = x_mat, z = z_mat, 
  timevar = gendat$time, indnum = gendat$indnum
)
```

### Fitting the ST Model
```{r fit1}
tic("AR(1)-ST-LMM Fit")
tbayes <- stan(file = 'lmm-AR1-ST.stan', 
               data = list(N=N, n=n, l=l_cols, q1=q1_cols, njvec=njvec, 
                           y=gendat$y, x=x_mat, z=z_mat, 
                           timevar=gendat$time, ind=gendat$indnum, sdLP=2), 
               thin = 5, chains = 3, iter = 5000, warmup = 1000, seed = 9955)
toc()

print(tbayes, pars=c("beta","sigmae","phi1","D1","lambda","nu"), probs=c(.025,.975), digits=3)

mcmc_dens_chains(tbayes, pars = c("beta[1]","beta[2]"))+
  geom_vline(xintercept = 0, linetype = "dotted") + theme_bw()
```

### Checking Acceptance Probability
```{r accept}
sampler_params <- get_sampler_params(tbayes, inc_warmup = FALSE)
mean(unlist(lapply(sampler_params, function(x) x[, "accept_stat__"])))
```

### Computing Highest (Posterior) Density Intervals (HDI)
```{r hdi}
tsample <- rstan::extract(tbayes, pars = c("beta[1]","beta[2]","sigmae","phi1","D1[1,1]",
                                    "D1[1,2]","D1[2,2]","lambda[1]","lambda[2]","nu"))
HDI<- HDInterval::hdi(tsample) %>% unlist()
data.frame(lower = HDI[c(TRUE, FALSE)],upper = HDI[c(FALSE, TRUE)]) %>% knitr::kable()
```

### Exploring the Posterior Distributions
```{r posteriors}
bayesplot::color_scheme_set("red")
bayesplot::mcmc_rank_hist(tbayes, pars=c("beta[1]","beta[2]","phi1","lambda[1]","lambda[2]","nu"))
```

Additional exploration can be easily done by launching the shinny application using `launch_shinystan(tbayes)`.

### Computing Information Criteria 

To evaluate the marginal log-likelihood function, an approximation of Leave-One-Out Cross-Validation (LOO), and the Widely applicable information criterion (WAIC) for all MCMC samples, we can use the function `criteriaAR1`:

```{r fit1criteria}
source("auxfunctions-bayes.R")

# ST Model LOO
st_loo <- criteriaAR1(tbayes, data_list_stan, distr = "st")

# Checking the Pareto smoothed importance sampling (PSIS)
plot(st_loo$loo)
```

## Model comparison with a simpler, Gaussian model

We compare the ST model against a standard Normal model.

```{r fit2}
# Normal Model Fitting
tic("AR(1)-N-LMM Fit")
nbayes <- stan(file = 'lmm-AR1-N.stan', 
               data = list(N=N, n=n, l=l_cols, q1=q1_cols, njvec=njvec, 
                           y=gendat$y, x=x_mat, z=z_mat, 
                           timevar=gendat$time, ind=gendat$indnum), 
               thin = 5, chains = 3, iter = 5000, warmup = 1000, seed = 9955)
toc()

# Computing the information criteria
normal_loo <- criteriaAR1(nbayes, data_list_stan, distr = "sn")

# Checking the Pareto smoothed importance sampling (PSIS)
plot(normal_loo$loo)
```

### Comparing the models based on LOO
```{r fit1loo}
comp_loo <- loo_compare(list(ST = st_loo$loo, Normal = normal_loo$loo))
knitr::kable(comp_loo, caption = "Model Comparison: LOO Criteria", simplify = FALSE)
```

### Comparing the models based on WAIC
```{r fit1waic}
comp_loo <- loo_compare(list(ST = st_loo$waic, Normal = normal_loo$waic))
knitr::kable(comp_loo, caption = "Model Comparison: WAIC Criteria", simplify = FALSE)
```

## Sensitivity analysis: changing prior distribution for shape parameter

To evaluate the model's sensitivity to the choice of the prior distribution for $\nu$, we now compare our original results based on a truncated Cauchy distribution with the ones considering a Gamma(2, 0.2) prior distribution for 
$\nu$.
```{r gammadist}
tic("AR(1)-ST-LMM Fit with Gamma prior")
tbayes_nu <- stan(file='lmm-AR1-ST-nuprior.stan', 
                  data = list(N=N, n=n, l=l_cols, q1=q1_cols, njvec=njvec, 
                           y=gendat$y, x=x_mat, z=z_mat, 
                           timevar=gendat$time, ind=gendat$indnum, sdLP=2), 
               thin = 5, chains = 3, iter = 5000, warmup = 1000, seed = 9955,
               control = list(adapt_delta=.95))
toc()
print(tbayes_nu,par=c("beta","sigmae","phi1","D1","lambda","nu"), 
      probs = c(.025,.975), digits=3)
```


```{r gammadistplot}
## Comparing with original model
posterior1 <- posterior::as_draws_matrix(tbayes)       
posterior2 <- posterior::as_draws_matrix(tbayes_nu)    

library(posterior)
d1 <- as_draws_df(posterior1) |>
  dplyr::select(nu, starts_with("beta")) |>
  tidyr::pivot_longer(cols = everything(), names_to = "param",values_to="values") |>
  dplyr::mutate(Model = "Cauchy Prior")
d2 <- as_draws_df(posterior2) |>
  dplyr::select(nu, starts_with("beta")) |>
  pivot_longer(cols = everything(), names_to = "param",values_to="values") |>
  dplyr::mutate(Model = "Gamma Prior")

d <- dplyr::bind_rows(d1, d2)

ggplot(d, aes(x = values, fill = Model)) +
  geom_density(alpha = 0.4) +
  ggtitle("Sensitivity of nu to Prior Choice") +
  theme_minimal() + facet_wrap(~param, scale = "free")
```


## Frequentist Comparison (EM Algorithm)

For validation, we compare our Bayesian estimates with the frequentist EM algorithm results.
```{r emfit}
tic("Frequentist Model")
fit_EM <- smsn.lmm(data = gendat, formFixed = y~x, groupVar = 'ind', 
                   formRandom = ~x, depStruct = "ARp", timeVar = 'time',
                   distr = 'st')
toc()
summary(fit_EM)

```
